---
title: "FinalPaper"
author: "John D Williams"
date: "2026-03-31"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Final Paper
```{r}
#Load Libraries
if (!require("tidyverse")) install.packages("tidyverse")
library(tidyverse)
if (!require("MASS")) install.packages("MASS")
library(MASS)
if (!require("ggplot2")) install.packages("ggplot2")
library(ggplot2)
if (!require("broom")) install.packages("broom")
library(broom)
```

## Read File 1
```{r}
df1 <- read_csv("CensusMCD.csv")
```

## Start cleaning process
```{r}
# 1. Convert Year to integer and Drop the Notes column
df1 <- df1 %>%
  mutate(Year = as.integer(Year))

# 2. Replace the text "Suppressed" with "5"  (the mean(Deaths))
df1$Deaths[df1$Deaths == "Suppressed"] <- "5"

# 3. Convert the entire Deaths column to numeric
df1$Deaths <- as.integer(df1$Deaths)

#Fix Race
# 4. We use mutate with ifelse (or case_when) to overwrite the values
df1 <- df1 %>%
  mutate(
    # Ensure both columns are strings to avoid type errors
    `Hispanic Origin` = as.character(`Hispanic Origin`),
    Race = as.character(Race),
    
    # Use case_when to precisely map the new Race values
    Race = case_when(
      `Hispanic Origin` == "Hispanic or Latino" ~ "Hispanic",
      `Hispanic Origin` == "Not Hispanic or Latino" ~ Race,
      is.na(`Hispanic Origin`) ~ Race, # Keeps original Race if Hispanic info is missing
      TRUE ~ Race                     # Safety net: keep original value otherwise
    ),
    
    # Do the same for the Codes
    `Race Code` = case_when(
      `Hispanic Origin` == "Hispanic or Latino" ~ `Hispanic Origin Code`,
      TRUE ~ as.character(`Race Code`)
    )
  )

# 5. Verification check
# This should now show "Hispanic", "White", "Black", AND "American Indian..."
unique(df1$Race)

# 6. Drop columns by name
df1_cleaned <- df1[, !(names(df1) %in% c("Hispanic Origin", "Hispanic Origin Code", "Notes"))]

# 7. Drop rows 2113 through 2200
df1_cleaned <- df1_cleaned %>%
  slice(-(2113:2200))

# See the count of each category
table(df1_cleaned$`Crude Rate`)

# See the count of each category
table(df1_cleaned$Deaths)

# See the count of each category
table(df1_cleaned$Race)

```

## Now do the same for all who died because of cirrosis

### Read File 2
```{r}
df2 <- read_csv("CensusOnlyK.csv")
```

## Start cleaning process for file 2
```{r}
# 1. Convert Year to integer and Drop the Notes column
df2 <- df2 %>%
  mutate(Year = as.integer(Year))

# 2. Replace the text "Suppressed" with "5"  (the mean(Deaths))
df2$Deaths[df2$Deaths == "Suppressed"] <- "5"

# 3. Convert the entire Deaths column to numeric
df2$Deaths <- as.integer(df2$Deaths)

#Fix Race
# 4. We use mutate with ifelse (or case_when) to overwrite the values
df2 <- df2 %>%
  mutate(
    # Ensure both columns are strings to avoid type errors
    `Hispanic Origin` = as.character(`Hispanic Origin`),
    Race = as.character(Race),
    
    # Use case_when to precisely map the new Race values
    Race = case_when(
      `Hispanic Origin` == "Hispanic or Latino" ~ "Hispanic",
      `Hispanic Origin` == "Not Hispanic or Latino" ~ Race,
      is.na(`Hispanic Origin`) ~ Race, # Keeps original Race if Hispanic info is missing
      TRUE ~ Race                     # Safety net: keep original value otherwise
    ),
    
    # Do the same for the Codes
    `Race Code` = case_when(
      `Hispanic Origin` == "Hispanic or Latino" ~ `Hispanic Origin Code`,
      TRUE ~ as.character(`Race Code`)
    )
  )

# 5. Verification check
# This should now show "Hispanic", "White", "Black", AND "American Indian..."
unique(df2$Race)

# 6. Drop columns by name
df2_cleaned <- df2[, !(names(df2) %in% c("Hispanic Origin", "Hispanic Origin Code", "Notes"))]

# 7. Drop rows 2113 through 2192
df2_cleaned <- df2_cleaned %>%
  slice(-(2113:2192))
```

## Fix Population 'Not Applicable' problem for df1 using values in df2
```{r}
# 1. Ensure Population is numeric (NAs will replace "Not Applicable")
df1_cleaned <- df1_cleaned %>%
  mutate(Population = as.numeric(gsub("[^0-9]", "", Population)))

df2_cleaned <- df2_cleaned %>%
  mutate(Population = as.numeric(gsub("[^0-9]", "", Population)))

# 2. Join the two datasets on your matching columns
# We use left_join to keep all 2112 rows of df1
df1_patched <- df1_cleaned %>%
  left_join(
    df2_cleaned %>% dplyr::select(Year, `Census Region`, Race, Sex, Population), 
    by = c("Year", "Census Region", "Race", "Sex"),
    suffix = c("_df1", "_df2")
  ) %>%
  # 3. Use coalesce to fill the holes
  mutate(Population = coalesce(Population_df1, Population_df2)) %>%
  # 4. Clean up the temporary columns
  dplyr::select(-Population_df1, -Population_df2)

# 5. Verify: How many NAs are left?
sum(is.na(df1_patched$Population))
```

## More df1 Cleanup
```{r}
# We use group_by to ensure we only fill within the same demographic bucket
df1_patched <- df1_patched %>%
  group_by(`Census Region`, Race, Sex) %>%
  arrange(Year) %>%
  # fill() looks at the nearest non-NA value and carries it forward or backward
  fill(Population, .direction = "downup") %>%
  ungroup()

# Check the count again
sum(is.na(df1_patched$Population))

#Calculate Your Final Rates
df1_final <- df1_patched %>%
  mutate(
    # Calculate the rate: (Deaths / Population) * 100,000
    Calculated_Crude_Rate = (Deaths / Population) * 100000
  )
```

## Fix df2_cleaned for the same items
```{r}
# 1. First, make sure Deaths in df2 are numeric (handling Suppressed/Unreliable)
df2_cleaned <- df2_cleaned %>%
  mutate(
    Deaths = case_when(
      Deaths == "Suppressed" ~ 5,
      Deaths == "Unreliable" ~ 15,
      TRUE ~ as.numeric(Deaths)
    )
  )

# 2. Join df2 with the "good" population from your patched df1
df2_final <- df2_cleaned %>%
  dplyr::select(-Population) %>% # Drop the old "dirty" population
  left_join(
    df1_final %>% dplyr::select(Year, `Census Region`, Race, Sex, Population), 
    by = c("Year", "Census Region", "Race", "Sex")
  )

# 3. Re-calculate the Crude Rate for df2
df2_final <- df2_final %>%
  mutate(Calculated_Crude_Rate = (Deaths / Population) * 100000)

# 4. Final Audit
sum(is.na(df2_final$Population))
```

## More Checking
```{r}
#The Final "Sanity Check"
df2_final %>%
  group_by(Race) %>%
  summarise(
    Avg_Crude_Rate_Total = mean(Calculated_Crude_Rate, na.rm = TRUE),
    Total_Deaths_Estimated = sum(Deaths, na.rm = TRUE)
  ) %>%
  arrange(desc(Avg_Crude_Rate_Total))


# Quick check: See the top rates by Race
df1_final %>%
  group_by(Race) %>%
  summarise(Avg_Crude_Rate = mean(Calculated_Crude_Rate)) %>%
  arrange(desc(Avg_Crude_Rate))

#A check of values
df2_final %>%
  group_by(Race) %>%
  summarise(Total_Pop = sum(Population, na.rm = TRUE))
```

## Aggregate Rates
```{r}
#The "Correct" Way to Calculate Aggregate Rates
Master_Summary <- df2_final %>%
  group_by(Race) %>%
  summarise(
    Grand_Total_Deaths = sum(Deaths, na.rm = TRUE),
    Grand_Total_Pop = sum(Population, na.rm = TRUE)
  ) %>%
  mutate(
    True_Crude_Rate = (Grand_Total_Deaths / Grand_Total_Pop) * 100000
  ) %>%
  arrange(desc(True_Crude_Rate))

print(Master_Summary)
```

## The Comparison: "Total" vs. "Both"
```{r}
# Calculate weighted rates for the comorbidity dataset
Both_Summary <- df1_final %>%
  group_by(Race) %>%
  summarise(
    Both_Deaths = sum(Deaths, na.rm = TRUE),
    Both_Pop = sum(Population, na.rm = TRUE)
  ) %>%
  mutate(
    Both_True_Rate = (Both_Deaths / Both_Pop) * 100000
  )

# Join them to see the final contribution
Final_Comparison <- Master_Summary %>%
  left_join(Both_Summary, by = "Race") %>%
  mutate(
    Diabetes_Contribution_Percent = (Both_Deaths / Grand_Total_Deaths) * 100
  )

print(Final_Comparison %>% dplyr::select(Race, True_Crude_Rate, 
  Both_True_Rate, Diabetes_Contribution_Percent))
```

## The Negative Binomial Regression
```{r}
# 0. Combine your two cleaned datasets into one master file
analysis_df <- df1_final %>%
  rename(Deaths_Both = Deaths) %>%
  # We select only the necessary columns from df2 to avoid 'doubling up' on population
  left_join(
    df2_final %>% dplyr::select(Year, `Census Region`, Race, Sex, Population),
    by = c("Year", "Census Region", "Race", "Sex", "Population")
  )

# Verify it exists now
head(analysis_df)
# 1. Ensure Race is a 'factor' and set 'White' as the baseline for comparison
analysis_df$Race <- factor(analysis_df$Race)
analysis_df$Race <- relevel(analysis_df$Race, ref = "White")

# 2. Run the Model
# We predict 'Deaths_Both' using 'Race' and 'Sex'
# We use log(Population) as an offset to calculate the RATE
nb_model <- glm.nb(Deaths_Both ~ Race + Sex + Year + offset(log(Population)), 
                   data = analysis_df)

# 3. See the results
summary(nb_model)
```

## Forest Plot
```{r}
# 1. Clean up the model results for plotting
plot_data <- tidy(nb_model, conf.int = TRUE, exponentiate = TRUE) %>%
  filter(term != "(Intercept)" & term != "Year") # Focus on Race and Sex

# 2. Create the Forest Plot - P1
ggplot(plot_data, aes(x = estimate, y = term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") + # The 'Null' line
  geom_point(size = 3, color = "darkblue") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
  theme_minimal() + 
  theme(plot.margin = margin(10, 100, 10, 10)) +
  labs(
    title = "Incidence Rate Ratios (IRR) for Cirrhosis-Diabetes Mortality",
    subtitle = "Reference Group: White | Adjusted for Year and Sex",
    x = "Risk Ratio (Exponential Coefficients)",
    y = "Demographic Variable"
  )
```

## Line Graph over Time
```{r}
# Create a line graph over time - P2
ggplot(df1_final, aes(x = Year, y = Calculated_Crude_Rate, color = Race)) +
  geom_smooth(se = FALSE, method = "loess") + # Smooths the lines to show the trend
  theme_minimal() +
  labs(
    title = "Mortality Trends: Cirrhosis and Type 2 Diabetes (1999-2020)",
    subtitle = "Aggregated by Race and Ethnicity",
    y = "Crude Rate per 100,000",
    x = "Year"
  )
```

## The Regional Risk Heatmap
```{r}
#Prep for the heatmap
regional_analysis <- analysis_df %>%
  group_by(`Census Region`, Race) %>%
  summarise(
    Total_Deaths = sum(Deaths_Both, na.rm = TRUE),
    # Using the calculated rate from df1_final
    Avg_Rate = mean(Calculated_Crude_Rate, na.rm = TRUE),
    .groups = 'drop'
  )

# Create the Heatmap - P3
ggplot(regional_analysis, aes(x = `Census Region`, y = Race, fill = Avg_Rate)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "firebrick") +
  theme_minimal() +
  theme(plot.margin = margin(10, 100, 10, 10)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Mortality Hotspots: Cirrhosis & Type 2 Diabetes",
    subtitle = "Average Crude Rate by Region and Race",
    fill = "Mortality Rate",
    x = "US Census Region",
    y = "Race/Ethnicity"
  )
```

## The Ratio Plot
```{r}
#One Last "New" Trick: The Ratio Plot - P4
analysis_df %>%
  group_by(`Census Region`, Race) %>%
  summarise(
    # What % of the Population is dying from this?
    Regional_Rate = (sum(Deaths_Both) / sum(Population)) * 100000,
    .groups = 'drop'
  ) %>%
  filter(Race %in% c("American Indian or Alaska Native", "White", "Hispanic", "Black or African American")) %>%
  ggplot(aes(x = `Census Region`, y = Regional_Rate, fill = Race)) +
  geom_bar(stat = "identity", position = "dodge") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Regional Mortality Rates by Race",
       y = "Deaths per 100,000",
       x = "Region")
```

## The "Inequality Ratio" by Region
```{r}
#Prep the data
regional_disparity <- analysis_df %>%
  group_by(`Census Region`, Race) %>%
  summarise(
    # Use the aggregate calculation (Sum of Deaths / Sum of Population)
    Agg_Rate = (sum(Deaths_Both) / sum(Population)) * 100000,
    .groups = 'drop'
  ) %>%
  # Pivot the data to compare Race vs White
  tidyr::pivot_wider(names_from = Race, values_from = Agg_Rate) %>%
  mutate(
    AIAN_Disparity_Ratio = `American Indian or Alaska Native` / White,
    Hispanic_Disparity_Ratio = Hispanic / White
  )

# View the ratios
print(regional_disparity %>% dplyr::select(`Census Region`,
 AIAN_Disparity_Ratio, Hispanic_Disparity_Ratio))
```

## The Final "Disparity Comparison" Plot
```{r}
# 1. Prepare the data for a side-by-side plot
plot_final <- regional_disparity %>%
  dplyr::select(`Census Region`, AIAN_Disparity_Ratio, Hispanic_Disparity_Ratio) %>%
  pivot_longer(cols = ends_with("Ratio"), 
               names_to = "Demographic_Group", 
               values_to = "Disparity_Ratio") %>%
  # Clean up labels for the legend
  mutate(Demographic_Group = ifelse(Demographic_Group == "AIAN_Disparity_Ratio", 
                                   "Amer. Indian / AK Native", "Hispanic"))

# 2. Create the Visual - P5
ggplot(plot_final, aes(x = `Census Region`, y = Disparity_Ratio, fill = Demographic_Group)) +
  geom_bar(stat = "identity", position = "dodge", color = "black", alpha = 0.8) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "red", size = 1) +
  annotate("text", x = 0.7, y = 1.3, label = "Baseline (White)", color = "red", fontface = "italic") +
  scale_fill_manual(values = c("Amer. Indian / AK Native" = "#b2182b", "Hispanic" = "#2166ac")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1, size = 10),
    legend.position = "top",
    plot.title = element_text(face = "bold", size = 14)
  ) +
  labs(
    title = "Regional Disparities in Cirrhosis-Diabetes Mortality",
    subtitle = "Relative Risk Compared to White Population (1.0 = Equality)",
    x = "US Census Region",
    y = "Disparity Ratio (Rate / White Rate)",
    fill = "Target Group"
  )
```

## Line Graph for Races
```{r}
# 1. Calculate the aggregate rate for each race per year
yearly_trends <- analysis_df %>%
  group_by(Year, Race) %>%
  summarise(
    Total_Deaths = sum(Deaths_Both, na.rm = TRUE),
    Total_Pop = sum(Population, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(Yearly_Rate = (Total_Deaths / Total_Pop) * 100000)

# 2. Plot the line graph - P6
ggplot(yearly_trends, aes(x = Year, y = Yearly_Rate, color = Race, group = Race)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Trends in Cirrhosis-Diabetes Mortality (1999–2020)",
    subtitle = "Crude Death Rates per 100,000 by Racial/Ethnic Group",
    x = "Year",
    y = "Death Rate (per 100k)",
    color = "Race/Ethnicity"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## The National Trend
```{r}
# 1. Aggregate everything by Year to get the national average
national_trend <- analysis_df %>%
  group_by(Year) %>%
  summarise(
    Total_Deaths = sum(Deaths_Both, na.rm = TRUE),
    Total_Pop = sum(Population, na.rm = TRUE),
    .groups = 'drop'
  ) %>%
  mutate(National_Rate = (Total_Deaths / Total_Pop) * 100000)

# 2. Create the "Surge" Line Graph - P7
ggplot(national_trend, aes(x = Year, y = National_Rate)) +
  geom_line(linewidth = 1.5, color = "darkred") +
  geom_point(size = 3, color = "darkred") +
  # Adding a trend line to show the "Slope of the Crisis"
  geom_smooth(method = "lm", linetype = "dashed", color = "gray40", se = FALSE) +
  theme_minimal() +
  labs(
    title = "The National Surge: Cirrhosis-Diabetes Comorbidity",
    subtitle = "Aggregated Death Rate per 100,000 (1999–2020)",
    x = "Year",
    y = "National Death Rate (per 100k)",
    caption = "Dashed line represents the overall linear trend."
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16),
    axis.title = element_text(face = "bold")
  )
```

## Sensitivity Analysis (The "Imputation" Check)
```{r}
# 1. Filter out the rows that were originally suppressed (where deaths were 5 or 15)
robust_df <- analysis_df %>% filter(Deaths_Both > 15)

# 2. Re-run the model
nb_model_robust <- glm.nb(Deaths_Both ~ Race + Sex + Year + offset(log(Population)), 
                          data = robust_df)

# 3. Compare the IRRs
summary(nb_model_robust)
```

## Predicted vs. Observed plot
```{r}
# Compare the model's predictions to the actual death counts - P8
analysis_df$predicted_deaths <- predict(nb_model, type = "response")

ggplot(analysis_df, aes(x = Deaths_Both, y = predicted_deaths)) +
  geom_point(alpha = 0.3, color = "darkblue") +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  theme_minimal() +
  labs(title = "Model Diagnostic: Predicted vs. Observed Deaths",
       x = "Actual Death Counts",
       y = "Model Predicted Counts")
```

## Prediction for 2026
```{r}
# First, let's check what data we actually have
# Create a clean trend plot with proper data
# Based on your code, I see you have yearly_trends data

# Create a clean national trend excluding 2020 outlier
cleaned_national_trend <- national_trend %>%
  filter(Year != 2020)  # Remove 2020 outlier

# Fit a proper linear model for trend
model <- lm(National_Rate ~ Year, data = cleaned_national_trend)
slope <- coef(model)[2]
intercept <- coef(model)[1]

# Predict 2026 rate
predicted_2026_rate <- intercept + slope * 2026

# Calculate percent increase from 2019 to 2026 (using 2019 as baseline)
actual_2019_rate <- national_trend %>% filter(Year == 2019) %>% pull(National_Rate)
percent_increase <- ((predicted_2026_rate - actual_2019_rate) / actual_2019_rate) * 100
```

## Print Results
```{r}
# Print results
cat("Predicted rate for 2026:", round(predicted_2026_rate, 2), "\n")
cat("Percent increase from 2019 to 2026:", round(percent_increase, 2), "%\n")
```

## Prep the data
```{r}
# Create the complete plot that shows the trend and prediction
# Get the actual data points
actual_data <- national_trend %>%
  mutate(Type = "Actual")

# Create future predictions
future_data <- data.frame(
  Year = 2021:2026,
  National_Rate = predict(model, newdata = data.frame(Year = 2021:2026)),
  Type = "Predicted"
)

# Combine all data
all_data <- bind_rows(actual_data, future_data)
```


## Create Initial Prediction graph
```{r}
# Create the final plot with proper x-axis
ggplot(all_data, aes(x = Year, y = National_Rate, color = Type)) +
  geom_line(size = 1.2, aes(linetype = Type)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 2020, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = 2026, linetype = "dashed", color = "gray50") +
  scale_linetype_manual(values = c("Actual" = "solid", "Predicted" = "dashed")) +
  theme_minimal() +
  labs(title = "National Trend in Cirrhosis-Diabetes Mortality (1999-2026)",
       subtitle = paste("Prediction for 2026:", round(predicted_2026_rate, 2), 
                       "per 100,000 | Increase:", round(percent_increase, 1), "%"),
       x = "Year",
       y = "National Death Rate (per 100,000)",
       color = "Data Type",
       linetype = "Data Type") +
  theme(plot.title = element_text(face = "bold", size = 16),
        plot.subtitle = element_text(size = 12),
        legend.position = "bottom")
```

## Line Plot Graph
```{r}
# Also create a simple line plot showing the trend
ggplot(national_trend, aes(x = Year, y = National_Rate)) +
  geom_line(size = 1.2, color = "darkblue") +
  geom_point(size = 3, color = "darkblue") +
  geom_line(data = data.frame(Year = 2021:2026, 
                              National_Rate = predict(model, newdata = data.frame(Year = 2021:2026))),
            aes(x = Year, y = National_Rate), 
            color = "red", linetype = "dashed", size = 1.2) +
  geom_point(data = data.frame(Year = 2021:2026, 
                               National_Rate = predict(model, newdata = data.frame(Year = 2021:2026))),
             aes(x = Year, y = National_Rate), 
             color = "red", size = 3) +
  theme_minimal() +
  labs(title = "National Trend in Cirrhosis-Diabetes Mortality (1999-2026)",
       subtitle = "Historical data with 2026 prediction",
       x = "Year",
       y = "National Death Rate (per 100,000)") +
  theme(plot.title = element_text(face = "bold", size = 16))
```

## Simpler Analysis
```{r}
# Simple approach using the last few years to predict
# Use 2018-2019 data to get trend
recent_data <- national_trend %>%
  filter(Year >= 2018 & Year <= 2020) %>%
  filter(Year != 2020)  # Remove 2020 outlier

# Fit model on recent data
model_recent <- lm(National_Rate ~ Year, data = recent_data)
slope_recent <- coef(model_recent)[2]
intercept_recent <- coef(model_recent)[1]

# Predict 2026
predicted_2026_rate_recent <- intercept_recent + slope_recent * 2026

# Calculate percent increase from 2019 to 2026
actual_2019_rate <- national_trend %>% filter(Year == 2019) %>% pull(National_Rate)
percent_increase_recent <- ((predicted_2026_rate_recent - actual_2019_rate) / actual_2019_rate) * 100

cat("Predicted rate for 2026 (using 2018-2019 trend):", round(predicted_2026_rate_recent, 2), "\n")
cat("Percent increase from 2019 to 2026:", round(percent_increase_recent, 2), "%\n")
```

